---
title: 'DATA CLEANING AND ANALYSIS: Survival Analysis of Time to 5% Body Weight Reduction Among Overweight Adults: A Longitudinal Study'
author: "MSc Group 1"
date: "2026-03-01"
output: 
  word_document:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. SETUP AND PACKAGE INSTALLATION

```{r setup-packages}
# Install required packages if needed
packages <- c("tidyverse", "survival", "survminer", "ggplot2", 
              "dplyr", "tidyr", "lubridate", "tableone", 
              "gridExtra", "ggfortify", "knitr")

# Install missing packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load all packages
lapply(packages, library, character.only = TRUE)

# Set global theme for plots
theme_set(theme_bw() + 
          theme(plot.title = element_text(hjust = 0.5, face = "bold"),
                legend.position = "bottom"))
```

# 2. DATA IMPORT AND PREPARATION

## 2.1 Import Raw Data

```{r import-data}
# Read the CSV file
df <- read.csv("C:\\Users\\munya\\Downloads\\MSc2026Group1_DATA_2026-03-01_1556.csv", 
               stringsAsFactors = FALSE)

# Display basic information
cat("Dataset dimensions:", dim(df)[1], "rows and", dim(df)[2], "columns\n")
cat("Column names:\n")
print(names(df))
```

## 2.2 Separate Baseline and Follow-up Data

```{r separate-data}
# Extract baseline data (enrollment visit)
baseline_raw <- df %>%
  filter(redcap_event_name == "baseline_arm_1") %>%
  select(
    id = participant_id,
    age,
    sex,
    date_enrolled,
    ethnic_group,
    province = what_province_are_you_from,
    baseline_weight,
    height_cm,
    bmi,
    target_weight = target_weight_5_reduction,
    have_you_lost_5_of_your_bo,
    date_of_birth
  ) %>%
  mutate(
    date_enrolled = as.Date(date_enrolled, format = "%d/%m/%Y"),
    date_of_birth = as.Date(date_of_birth, format = "%d/%m/%Y"),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female"))
  )

# Extract follow-up data (monthly visits)
followup_raw <- df %>%
  filter(redcap_event_name %in% c("month_1_arm_1", "month_2_arm_1", "month_3_arm_1", 
                                   "month_4_arm_1", "month_5_arm_1", "month_6_arm_1")) %>%
  select(
    id = participant_id,
    event = redcap_event_name,
    visit_date,
    current_weight,
    pct_loss = percent_weight_loss,
    event_status = event_indicator,
    phys_activity = any_physical_activity,
    phys_days = physical_activity_days,
    diet = on_a_diet,
    diet_score = diet_quality,
    fastfood = ave_fastfood_consumption,
    type_1 = type_of_physical_activity___1,
    type_2 = type_of_physical_activity___2,
    type_3 = type_of_physical_activity___3,
    type_4 = type_of_physical_activity___4,
    type_5 = type_of_physical_activity___5
  ) %>%
  mutate(
    visit_date = as.Date(visit_date, format = "%d/%m/%Y"),
    # Create physical activity type
    phys_type = case_when(
      type_2 == 1 ~ "Running",
      type_3 == 1 ~ "Swimming",
      type_4 == 1 ~ "Gym",
      type_1 == 1 ~ "Walking",
      type_5 == 1 ~ "Other",
      TRUE ~ "None"
    )
  )
```

## 2.3 Merge Baseline and Follow-up Data

```{r merge-data}
# Merge datasets
clean_data <- followup_raw %>%
  left_join(baseline_raw, by = "id") %>%
  mutate(
    # Extract month number from event
    month_num = case_when(
      event == "month_1_arm_1" ~ 1,
      event == "month_2_arm_1" ~ 2,
      event == "month_3_arm_1" ~ 3,
      event == "month_4_arm_1" ~ 4,
      event == "month_5_arm_1" ~ 5,
      event == "month_6_arm_1" ~ 6,
      TRUE ~ NA_real_
    ),
    # Create physical activity level categories
    phys_level = case_when(
      phys_days >= 5 ~ "High (5+ days)",
      phys_days >= 3 ~ "Moderate (3-4 days)",
      phys_days >= 1 ~ "Low (1-2 days)",
      phys_days == 0 ~ "Sedentary",
      is.na(phys_days) & phys_activity == 0 ~ "Sedentary",
      TRUE ~ "Unknown"
    ),
    # Create diet quality categories
    diet_cat = case_when(
      diet_score >= 7 ~ "Good (7-10)",
      diet_score >= 4 ~ "Fair (4-6)",
      diet_score >= 1 ~ "Poor (1-3)",
      TRUE ~ "Unknown"
    ),
    # Fast food categories
    fastfood_cat = case_when(
      fastfood == 0 ~ "Never",
      fastfood == 1 ~ "Rarely",
      fastfood == 2 ~ "Sometimes",
      fastfood == 3 ~ "Often",
      fastfood >= 4 ~ "Very Often",
      TRUE ~ "Unknown"
    )
  ) %>%
  # Remove rows with no weight data
  filter(!is.na(current_weight) | !is.na(pct_loss))
```

## 2.4 Verify Data Structure

```{r verify-data}
# Check merged data
cat("=== MERGED DATA SUMMARY ===\n")
cat("Total rows:", nrow(clean_data), "\n")
cat("Unique participants:", length(unique(clean_data$id)), "\n")
cat("Age range:", range(clean_data$age, na.rm = TRUE), "\n")
cat("\nSex distribution:\n")
print(table(clean_data$sex, useNA = "ifany"))

# View sample of data
clean_data %>%
  select(id, event, month_num, age, sex, baseline_weight, current_weight, pct_loss, event_status) %>%
  head(10) %>%
  knitr::kable(caption = "Sample of Clean Data")
```

# 3. CREATE SURVIVAL ANALYSIS DATASET

## 3.1 Calculate Time Variables

```{r time-variables}
# Calculate time from enrollment to each visit
time_to_event <- clean_data %>%
  mutate(
    days_since_enrollment = as.numeric(difftime(visit_date, date_enrolled, units = "days")),
    weeks_since_enrollment = days_since_enrollment / 7,
    months_since_enrollment = month_num
  )
```

## 3.2 Build Survival Dataset

```{r survival-dataset}
# Create one row per participant with survival information
survival_data <- time_to_event %>%
  group_by(id) %>%
  summarise(
    # Time to event (if achieved)
    time_to_event_days = min(days_since_enrollment[event_status == 1], na.rm = TRUE),
    time_to_event_weeks = min(weeks_since_enrollment[event_status == 1], na.rm = TRUE),
    time_to_event_months = min(months_since_enrollment[event_status == 1], na.rm = TRUE),
    
    # Event occurrence
    event_occurred = ifelse(any(event_status == 1, na.rm = TRUE), 1, 0),
    
    # Last follow-up time (for censored)
    last_followup_days = max(days_since_enrollment, na.rm = TRUE),
    last_followup_weeks = max(weeks_since_enrollment, na.rm = TRUE),
    last_followup_months = max(months_since_enrollment, na.rm = TRUE),
    
    # Baseline characteristics
    age = first(na.omit(age)),
    sex = first(na.omit(sex)),
    bmi = first(na.omit(bmi)),
    baseline_weight = first(na.omit(baseline_weight)),
    
    .groups = "drop"
  ) %>%
  mutate(
    # Determine final time (event time for successes, last follow-up for censored)
    time_days = ifelse(event_occurred == 1, time_to_event_days, last_followup_days),
    time_weeks = ifelse(event_occurred == 1, time_to_event_weeks, last_followup_weeks),
    time_months = ifelse(event_occurred == 1, time_to_event_months, last_followup_months),
    
    # Handle infinite/NA values
    time_days = ifelse(is.infinite(time_days) | is.na(time_days), 0, time_days),
    time_weeks = ifelse(is.infinite(time_weeks) | is.na(time_weeks), 0, time_weeks),
    time_months = ifelse(is.infinite(time_months) | is.na(time_months), 0, time_months),
    
    # Create BMI categories
    bmi_cat = case_when(
      bmi < 25 ~ "Normal",
      bmi < 30 ~ "Overweight",
      bmi >= 30 ~ "Obese",
      TRUE ~ "Unknown"
    )
  )
```

# 4. DESCRIPTIVE STATISTICS

## 4.1 Study Population Summary

```{r summary-stats}
cat("\n=== STUDY POPULATION SUMMARY ===\n")
cat("Total participants:", nrow(survival_data), "\n")
cat("Achieved 5% weight loss:", sum(survival_data$event_occurred), 
    sprintf("(%.1f%%)", mean(survival_data$event_occurred) * 100), "\n")

cat("\nSex distribution:\n")
print(table(survival_data$sex, useNA = "ifany"))

cat("\nBMI categories:\n")
print(table(survival_data$bmi_cat, useNA = "ifany"))

cat("\nAge summary:\n")
print(summary(survival_data$age))

cat("\nBMI summary:\n")
print(summary(survival_data$bmi))
```

## 4.2 Table 1: Participant Characteristics

```{r table1}
# Create Table 1 stratified by event occurrence
vars <- c("age", "sex", "bmi")
factor_vars <- c("sex")

table1 <- CreateTableOne(vars = vars, 
                         strata = "event_occurred", 
                         data = survival_data,
                         factorVars = factor_vars)

print(table1, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
```

# 5. KAPLAN-MEIER SURVIVAL ANALYSIS

## 5.1 Overall Survival

```{r km-overall}
# Create survival object
surv_obj <- Surv(time = survival_data$time_weeks, 
                 event = survival_data$event_occurred)

# Overall Kaplan-Meier fit
km_overall <- survfit(surv_obj ~ 1, data = survival_data)

# Print results
cat("\n=== OVERALL SURVIVAL ANALYSIS ===\n")
print(km_overall)

# Plot overall survival
plot_overall <- ggsurvplot(
  km_overall,
  data = survival_data,
  title = "Overall Probability of Not Achieving 5% Weight Loss",
  xlab = "Time (weeks)",
  ylab = "Survival Probability",
  legend = "none",
  conf.int = TRUE,
  risk.table = TRUE,
  risk.table.height = 0.25,
  ggtheme = theme_bw(),
  break.time.by = 4
)

print(plot_overall)
```

## 5.2 Survival by Sex

```{r km-sex}
# KM by sex
km_sex <- survfit(surv_obj ~ sex, data = survival_data)

# Plot by sex
plot_sex <- ggsurvplot(
  km_sex,
  data = survival_data,
  title = "Time to 5% Weight Loss by Sex",
  xlab = "Time (weeks)",
  ylab = "Survival Probability",
  legend.title = "Sex",
  conf.int = TRUE,
  risk.table = TRUE,
  risk.table.height = 0.25,
  ggtheme = theme_bw(),
  break.time.by = 4,
  pval = TRUE
)

print(plot_sex)

# Log-rank test
cat("\nLog-rank test for sex difference:\n")
print(survdiff(surv_obj ~ sex, data = survival_data))
```

## 5.3 Survival by BMI Category

```{r km-bmi}
# KM by BMI category
km_bmi <- survfit(surv_obj ~ bmi_cat, data = survival_data)

# Plot by BMI
plot_bmi <- ggsurvplot(
  km_bmi,
  data = survival_data,
  title = "Time to 5% Weight Loss by BMI Category",
  xlab = "Time (weeks)",
  ylab = "Survival Probability",
  legend.title = "BMI Category",
  conf.int = TRUE,
  risk.table = TRUE,
  risk.table.height = 0.25,
  ggtheme = theme_bw(),
  break.time.by = 4,
  pval = TRUE
)

print(plot_bmi)

# Log-rank test
cat("\nLog-rank test for BMI category difference:\n")
print(survdiff(surv_obj ~ bmi_cat, data = survival_data))
```

## 5.4 Create Physical Activity Categories

```{r phys-categories}
# Add physical activity categories from follow-up data
phys_summary <- time_to_event %>%
  group_by(id) %>%
  summarise(avg_phys_days = mean(phys_days, na.rm = TRUE)) %>%
  mutate(
    phys_cat = case_when(
      avg_phys_days <= 2 ~ "Low (0-2 days/week)",
      avg_phys_days <= 4 ~ "Moderate (3-4 days/week)",
      avg_phys_days >= 5 ~ "High (5+ days/week)",
      TRUE ~ "Unknown"
    )
  )

survival_data <- survival_data %>%
  left_join(phys_summary %>% select(id, phys_cat), by = "id")
```

## 5.5 Survival by Physical Activity Level

```{r km-phys}
# KM by physical activity
km_phys <- survfit(surv_obj ~ phys_cat, data = survival_data)

# Plot by physical activity
plot_phys <- ggsurvplot(
  km_phys,
  data = survival_data,
  title = "Time to 5% Weight Loss by Physical Activity Level",
  xlab = "Time (weeks)",
  ylab = "Survival Probability",
  legend.title = "Physical Activity",
  conf.int = TRUE,
  risk.table = TRUE,
  risk.table.height = 0.25,
  ggtheme = theme_bw(),
  break.time.by = 4,
  pval = TRUE
)

print(plot_phys)

# Log-rank test
cat("\nLog-rank test for physical activity difference:\n")
print(survdiff(surv_obj ~ phys_cat, data = survival_data))
```

## 5.6 Create Diet Quality Categories

```{r diet-categories}
# Add diet categories from follow-up data
diet_summary <- time_to_event %>%
  group_by(id) %>%
  summarise(avg_diet_score = mean(diet_score, na.rm = TRUE)) %>%
  mutate(
    diet_cat = case_when(
      avg_diet_score <= 3 ~ "Poor (1-3)",
      avg_diet_score <= 6 ~ "Fair (4-6)",
      avg_diet_score >= 7 ~ "Good (7-10)",
      TRUE ~ "Unknown"
    )
  )

survival_data <- survival_data %>%
  left_join(diet_summary %>% select(id, diet_cat), by = "id")
```

## 5.7 Survival by Diet Quality

```{r km-diet}
# KM by diet quality
km_diet <- survfit(surv_obj ~ diet_cat, data = survival_data)

# Plot by diet quality
plot_diet <- ggsurvplot(
  km_diet,
  data = survival_data,
  title = "Time to 5% Weight Loss by Diet Quality",
  xlab = "Time (weeks)",
  ylab = "Survival Probability",
  legend.title = "Diet Quality",
  conf.int = TRUE,
  risk.table = TRUE,
  risk.table.height = 0.25,
  ggtheme = theme_bw(),
  break.time.by = 4,
  pval = TRUE
)

print(plot_diet)

# Log-rank test
cat("\nLog-rank test for diet quality difference:\n")
print(survdiff(surv_obj ~ diet_cat, data = survival_data))
```

# 6. COX PROPORTIONAL HAZARDS REGRESSION

## 6.1 Univariate Cox Models

```{r cox-univariate}
# Run univariate Cox models
cox_sex <- coxph(surv_obj ~ sex, data = survival_data)
cox_age <- coxph(surv_obj ~ age, data = survival_data)
cox_bmi <- coxph(surv_obj ~ bmi, data = survival_data)

# Display results
cat("\n=== SEX (Univariate) ===\n")
print(summary(cox_sex))

cat("\n=== AGE (Univariate) ===\n")
print(summary(cox_age))

cat("\n=== BMI (Univariate) ===\n")
print(summary(cox_bmi))
```

## 6.2 Multivariable Cox Model

```{r cox-multivariable}
# Multivariable Cox model
cox_multi <- coxph(surv_obj ~ sex + age + bmi, data = survival_data)

cat("\n=== MULTIVARIABLE COX MODEL ===\n")
print(summary(cox_multi))

# Create hazard ratio table
hr_table <- data.frame(
  Variable = c("Sex (Female vs Male)", "Age (per year)", "BMI (per unit)"),
  HR = round(exp(coef(cox_multi)), 2),
  CI_lower = round(exp(confint(cox_multi))[,1], 2),
  CI_upper = round(exp(confint(cox_multi))[,2], 2),
  P_value = round(summary(cox_multi)$coefficients[,5], 3)
)

hr_table$CI <- paste0(hr_table$CI_lower, "-", hr_table$CI_upper)
hr_table <- hr_table %>% select(Variable, HR, CI, P_value)

knitr::kable(hr_table, caption = "Multivariable Cox Model Results")
```

## 6.3 Forest Plot of Hazard Ratios

```{r forest-plot}
# Create forest plot
hr_data <- data.frame(
  variable = c("Sex (Female vs Male)", "Age (per year)", "BMI (per unit)"),
  HR = exp(coef(cox_multi)),
  lower = exp(confint(cox_multi)[,1]),
  upper = exp(confint(cox_multi)[,2])
)

hr_data$variable <- factor(hr_data$variable, levels = rev(hr_data$variable))

forest_plot <- ggplot(hr_data, aes(x = HR, y = variable)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2, size = 0.8) +
  geom_point(size = 4, color = "steelblue") +
  scale_x_log10(breaks = c(0.5, 0.75, 1, 1.25, 1.5)) +
  labs(title = "Predictors of Time to 5% Weight Loss",
       x = "Hazard Ratio (95% CI)",
       y = "") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        panel.grid.minor = element_blank())

print(forest_plot)
```

## 6.4 Check Proportional Hazards Assumption

```{r proportional-hazards}
# Test proportional hazards assumption
cox_zph <- cox.zph(cox_multi)
print(cox_zph)

# Plot Schoenfeld residuals
par(mfrow = c(2, 2))
plot(cox_zph)
```

# 7. WEIGHT LOSS TRAJECTORIES

## 7.1 Individual Participant Trajectories

```{r trajectories}
# Prepare trajectory data
traj_data <- time_to_event %>%
  filter(!is.na(pct_loss)) %>%
  mutate(
    achieved = ifelse(id %in% survival_data$id[survival_data$event_occurred == 1], 
                      "Achieved 5%", "Did not achieve")
  )

# Plot trajectories
traj_plot <- ggplot(traj_data, aes(x = weeks_since_enrollment, 
                                    y = pct_loss, 
                                    group = id,
                                    color = achieved)) +
  geom_line(alpha = 0.6, size = 0.8) +
  geom_hline(yintercept = -5, linetype = "dashed", color = "red", size = 1) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray", size = 0.5) +
  scale_color_manual(values = c("Achieved 5%" = "darkgreen", 
                                 "Did not achieve" = "gray70")) +
  labs(title = "Individual Weight Loss Trajectories",
       x = "Time (weeks)",
       y = "Percent Weight Loss (%)",
       color = "") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom")

print(traj_plot)
```

## 7.2 Cumulative Incidence

```{r cumulative-incidence}
# Calculate cumulative incidence
cum_inc <- data.frame(
  time = summary(km_overall)$time,
  surv = summary(km_overall)$surv,
  cum_inc = 1 - summary(km_overall)$surv,
  lower = 1 - summary(km_overall)$upper,
  upper = 1 - summary(km_overall)$lower
)

# Plot cumulative incidence
cum_plot <- ggplot(cum_inc, aes(x = time, y = cum_inc)) +
  geom_step(size = 1.2, color = "darkblue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              alpha = 0.2, fill = "blue") +
  labs(title = "Cumulative Incidence of 5% Weight Loss",
       x = "Time (weeks)",
       y = "Proportion Achieving Goal") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(cum_plot)
```

# 8. RISK TABLES AND KEY FINDINGS

## 8.1 Risk Table at Key Time Points

```{r risk-table}
# Extract survival probabilities at key time points
time_points <- c(4, 8, 12, 16, 20, 24)
surv_summary <- summary(km_overall, times = time_points)

risk_table <- data.frame(
  `Time (weeks)` = time_points,
  `Number at Risk` = surv_summary$n.risk,
  `Events` = surv_summary$n.event,
  `Survival Probability` = round(surv_summary$surv, 3),
  `Cumulative Incidence` = round(1 - surv_summary$surv, 3)
)

knitr::kable(risk_table, caption = "Risk Table at Key Time Points", digits = 3)
```

## 8.2 Summary Statistics for Report

```{r final-summary}
# Calculate key statistics
summary_stats <- list(
  n_total = nrow(survival_data),
  n_events = sum(survival_data$event_occurred),
  event_percent = mean(survival_data$event_occurred) * 100,
  mean_age = mean(survival_data$age, na.rm = TRUE),
  sd_age = sd(survival_data$age, na.rm = TRUE),
  percent_female = mean(survival_data$sex == "Female", na.rm = TRUE) * 100,
  mean_bmi = mean(survival_data$bmi, na.rm = TRUE),
  sd_bmi = sd(survival_data$bmi, na.rm = TRUE),
  median_survival_weeks = summary(km_overall)$table["median"]
)

# Print formatted summary
cat("\n", rep("=", 50), "\n", sep="")
cat("           STUDY SUMMARY STATISTICS\n")
cat(rep("=", 50), "\n\n", sep="")
cat(sprintf("Total participants: %d\n", summary_stats$n_total))
cat(sprintf("Achieved 5%% weight loss: %d (%.1f%%)\n", 
            summary_stats$n_events, summary_stats$event_percent))
cat(sprintf("Mean age: %.1f years (SD = %.1f)\n", 
            summary_stats$mean_age, summary_stats$sd_age))
cat(sprintf("Female participants: %.1f%%\n", summary_stats$percent_female))
cat(sprintf("Mean BMI: %.1f kg/m² (SD = %.1f)\n", 
            summary_stats$mean_bmi, summary_stats$sd_bmi))
cat(sprintf("Median time to 5%% weight loss: %.1f weeks\n", 
            summary_stats$median_survival_weeks))
```

# 9. EXPORT RESULTS

## 9.1 Save Plots

```{r save-plots}
# Create figures directory
if(!dir.exists("figures")) dir.create("figures")

# Save all plots
ggsave("figures/01_overall_survival.png", plot_overall$plot, 
       width = 8, height = 6, dpi = 300)
ggsave("figures/02_survival_by_sex.png", plot_sex$plot, 
       width = 8, height = 6, dpi = 300)
ggsave("figures/03_survival_by_bmi.png", plot_bmi$plot, 
       width = 8, height = 6, dpi = 300)
ggsave("figures/04_survival_by_phys_activity.png", plot_phys$plot, 
       width = 8, height = 6, dpi = 300)
ggsave("figures/05_survival_by_diet.png", plot_diet$plot, 
       width = 8, height = 6, dpi = 300)
ggsave("figures/06_forest_plot.png", forest_plot, 
       width = 8, height = 5, dpi = 300)
ggsave("figures/07_weight_trajectories.png", traj_plot, 
       width = 10, height = 6, dpi = 300)
ggsave("figures/08_cumulative_incidence.png", cum_plot, 
       width = 8, height = 6, dpi = 300)

cat("\n✅ All plots saved to 'figures' folder\n")
```

## 9.2 Export Data

```{r export-data}
# Export datasets
write.csv(survival_data, "survival_analysis_data.csv", row.names = FALSE)
write.csv(hr_table, "cox_model_results.csv", row.names = FALSE)
write.csv(risk_table, "risk_table.csv", row.names = FALSE)

cat("\n✅ Data exported to CSV files\n")
```

## 9.3 Session Information

```{r session-info}
# Save session info for reproducibility
sink("session_info.txt")
sessionInfo()
sink()

cat("\n✅ Session info saved to 'session_info.txt'\n")
```

---

```{r echo=FALSE, results='asis'}
cat("\n\n---\n")
cat("**Analysis complete!** Generated on: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
cat("**Total files created:** 8 plots, 3 CSV files, 1 session info file\n")
```
